\singlespacing{\textbf{Efficient pollution abatement in electricity markets with intermittent renewable energy}}

### 2019-05-01 documents/draft.tex
Update draft
- clean up introduction
- clarify methodology
- fix results formatting

\section{Extended Abstract}

Renewable energy technologies have seen considerable adoption over the last few decades (EIA 2019). Unlike the alternatives, wind and solar power are particularly unique in that the amount of energy they supply is intermittent. Consequently, designing economically efficient policy to promote their adoption is not straightforward given that they cannot easily substitute for traditional technologies such as coal power which handles base load demand. Some of the literature has approached the problem by constructing numerical models that find the cheapest renewable technology set while accounting for intermittent supply. Müsgens and Neuhoff (2006) model uncertain renewable output with intertemporal generation constraints, while Neuhoff, Cust, and Keats (2007) model temporal and spatial characteristics of wind output to optimize its deployment in the UK. On the other hand, other literature focuses on the effect of intermittent technologies on the market itself; Ambec and Crampes (2010) study the interaction between intermittent renewables and traditional reliable sources of energy in decentralized markets, and Chao (2011) models alternative pricing mechanisms for intermittent renewable energy sources. Additionally, Borenstein (2010) reviews the effects of present public policies used to promote renewables and the challenges posed by intermittency.

In this paper, we present a theoretical model of electricity markets with intermittent renewable energy and derive the optimal public policy to handle pollution externalities. In contrast with other theoretical models of intermittency which optimize a pubic electricity sector, our model assumes utility and profit maximization. Additionally, it represents the energy sector over multiple periods with electricity output for each technology varying over time for intermittent technologies. This approach is better suited for studying the dynamics of present US electricity markets which are primarily funded by private sector investment and have prices that vary over time. We first consider a simpler version of our model with two periods and two energy generation technologies, derive the comparative statics for this model, and detail the policy implications. Then, we produce numerical results for optimal policy prescriptions with a multi-period, multi-technology version of our model using electricity generation data on the PGM region.

Present models of the energy sector concerned with the adoption of renewable energy often discuss the elasticity of substitution between clean and dirty energy; this elasticity is meant to capture factors such as intermittency and reliability that impede perfect substitution between energy sources. Additionally, this elasticity is often estimated empirically or modelled in a CGE using a CES production function with capital inputs for each energy technology. However, this top down approach may be missing the actual source of the substitution effects while also being unrealistic. For instance, consider a simple case where energy output is a Cobb-Douglas function (a special case of CES) of two energy inputs: solar and coal. According to this function, increasing the amount of coal input causes the marginal product of solar input to rise; this makes little sense in practical terms, since the output of additional solar panels should not be related to the amount of coal input. Moreover, this approach, while possibly relevant for other sectors' goods, weakens the accuracy of theoretical and CGE models focused on the energy sector.

However, it is still possible to produce a more accurate model of the electricity sector that can capture trade-offs between the energy output of different technologies from both cost and intermittency. To start, rather than energy production following a general CES function, we first assume that energy follows a linear production function where total energy is instead the sum of the energy output of each source. The purpose of using a linear production function is that increasing the input of one energy source does not change the marginal product of another source as it does in a CES function. Additionally, we split this production function up across periods; that is, in each period, the electricity generation is equal to the energy output of each source in that period. For intermittent technologies, energy output varies in each period, so total output would as well. Next, we assume that firms pick a profit-maximizing level of investment in each energy technology. This investment stays fixed in all periods, but the total energy output can vary over time due to intermittency; for instance, a solar plant provides a variable output by time of day, while nuclear plant cannot easily vary its output within 24 hours. Moreover, we assume that firms face linear costs and electricity prices are set equal to their marginal cost; the former assumption is temporarily made for mathematical tractability while the latter is realistic given the competitiveness of electricity markets.

While the production function described so far is linear, we may still find a non-zero elasticity of substitution between energy sources. That is, consider a representative agent with a CES utility function that captures intertemporal variation in the utility gained from energy consumption. Specifically, the utility function is composed of energy consumption differentiated by period; so, for example, in a two-period model, we may have off-peak consumption and on-peak consumption as our two goods. Because people prefer to consume energy at different proportions based on the time of day, this variation can be modeled through the share parameters in the CES function. Moreover, since people may substitute energy consumption across time by availability and prices, the CES elasticity parameter captures the intertemporal elasticity of substitution. Thus, in this model, the intermittency of an energy generation technology plays a key role in determining its substitutability with other technologies. For instance, solar power may be a good substitute for coal power during the day, but obviously not at night; consequently, this pair complements each other, since coal handles the base load while solar handles the peak load. Alternatively, wind and solar are a poor combination, since both produce intermittently; this pair is closer to being substitutes. All in all, the result of using CES preferences with temporally differentiated energy consumption is that we can capture substitutability/complementarity between energy sources in an accurate way.

The first significant result in this model is that the optimal quantity of investment in intermittent renewable technology is concave with respect to both cost efficiency (cost per unit) and output efficiency (energy output per unit). So, for example, suppose some location has a coal-fired plant and is considering investing in a solar power plant. A 10\% increase in the output per solar panel may increase the optimal quantity of solar panels by x\%, while a 20\% increase in solar efficiency will increase the optimal quantity of solar by y\% where y < 2x. Because of this concavity, exponential increases in efficiency of intermittent renewable sources over time, as Moore’s Law may predict, is not enough to fully substitute out reliable energy technology. Hence, we argue that a full transition to renewable energy requires more emphasis on technologies such as nuclear, biomass, hydro, and geothermal energy sources; although these sources compete with intermittent energy technology, they are able fully substitute out for fossil fuel energy. Additionally, the reliable output of these technologies can handle base loads; this allows them to complement the intermittency of renewables such as wind and solar.

Secondly, suppose that we would like to promote clean energy adoption and replace dirty energy to reduce the pollution emissions. Because we have diminishing returns from both forms of efficiency, it is optimal to subsidize both research and the cost of intermittent renewable technology. This is because we expect research subsidies to increase both output efficiency and cost efficiency, while subsidies would increase the cost efficiency of renewables for a market participant. A mix of both instruments leads to improving both forms of efficiency at once, thus replacing dirty energy technology in the most cost-efficient way. Alternatively, since we would see symmetrical effects from taxing dirty energy sources rather than subsidizing clean ones, taxing dirty energy sources can substitute for renewable cost subsidies. Consequently, a carbon tax plus a research subsidy for renewables is another optimal choice for pollution abatement.

Finally, our model implies that the optimal level of subsidies/taxes on energy generation should vary by the state of the local energy market. That is, the change in optimal quantity of intermittent energy technology with respect to both types of efficiency is also a function of the efficiencies of the other technologies available. Consequently, the marginal benefit of a subsidy or tax varies spatially, since different geographies have access to different energy sources. For instance, one community may be relying on hydropower while another may be using coal power; the effect of a 1\% cost subsidy on solar installments in these two communities would vary because of the differences in the communities’ pre-existing energy generation technologies. Hence, local communities that aim to promote clean energy and reduce pollution should optimize their policy instruments to suit their local energy markets.

\subsection{References}

\begin{itemize}
	\item Literature
	\item Description of model
	\item 
\end{itemize}

\section{Outline}
\dots

Things I'd like to say:
\begin{itemize}
	\item Because the marginal product of renewable energy inputs may be time-dependent, renewable energy may complement traditional, constant-output energy sources rather than fully substitute for it. Obvious arugment follows from base load output required at all times. 
	\item How the conversion parameters for energy inputs affect their elasticity of substitution
	\item How variability in energy output decreases the value of particular energy sources. Obvious argument from concavity of utility implying risk aversion. 
\end{itemize}

@@

Renewable energy technologies have seen considerable adoption over the last few decades (EIA 2019). Unlike the alternatives, wind and solar power are particularly unique in that the amount of energy they supply is intermittent. Consequently, designing economically efficient policy to promote their adoption is not straightforward given that they cannot easily substitute for traditional technologies such as coal power which handles base load demand. Moreover, simpler approaches, such as using the levelized cost of electricity (LCOE) to compare technologies is flawed in any setting with variable electricity prices. In particular, \citet{Joskow2011} argues that

\begin{quote}
	``Instead [of using metrics like LCOE], the economics of all generating technologies, both intermittent and dispatchable, can be evaluated based on the expected market value of the electricity that
	they will supply, their total life-cycle costs, and their associated expected profitability. Such an
	analysis would reflect the actual expected production profiles of dispatchable and intermittent
	technologies, the value of electricity supplied at different times, and other costs of intermittency
	associated with reliable network integration. This is exactly the way investors in merchant generating plants evaluate whether or not to invest."
\end{quote}
We structure our model with this critique in mind. Specifically, we develop a model where producers optimize the supply of each generation technology on the expected market value of the electricity produce. We do so by ....

Some of the literature has approached the problem by constructing numerical models that find the cheapest renewable technology set while accounting for intermittent supply. \citet{MN2006} model uncertain renewable output with intertemporal generation constraints, while \citet{NCK2007} model temporal and spatial characteristics of wind output to optimize its deployment in the UK. On the other hand, other literature focuses on the effect of intermittent technologies on the market itself; \citet{AC2012} study the interaction between intermittent renewables and traditional reliable sources of energy in decentralized markets, and \citet{Chao2011} models alternative pricing mechanisms for intermittent renewable energy sources. Additionally, \citet{Boren2012} reviews the effects of present public policies used to promote renewables and the challenges posed by intermittency.

Our model comes closest to that of \citet{HH} who build a peak-load pricing model where the availability of renewable capacity varies stochastically. This stochastic variability models the intermittency and affects the adoption of renewables in a market with dynamic pricing. Furthermore, they simplify their analysis by setting aside complications such as  outage costs and rationing rules. In total, their model finds an S-shaped adoption curve for renewables once their LCOE's reach those of fossil fuel energy. In addition, they find that a Pigouvian tax can properly internalize the costs of fossil fuels in a setting with perfect competition. Likewise, our model makes equivalent simplifications and finds similar results on an equilibrium in  amarket with dynamic prices. But, our representation of the intermittency of renewables differs in a significant way.

To highlight this difference, we must first contrast the terms  \textit{intermittency} and \textit{reliability}. By intermittency, we predictable changes in output related to physical constraints. For instance, while wind energy output varies over time, if  we know the angle and speed of wind, we may precisely derive the quantity of energy that a wind farm generates.\footnote{\citet{Foley2012} provide a review of the literature on the forecasting of wind power. Over time, forecasts have gotten much more accurate allowing electricity grids to manage wind power intermittency ahead of time.}  At the same time, these forecasts may not be perfect and, as a result, we see unexpected variation in the output of a technology. This unpredictability may be better captured by the notion of reliability; specifically,  we refer to definition provided by the US Department of Energy's \citep{ORNL} -- ``Power reliability can be defined as the degree to which the performance of the elements in a bulk system results in electricity being delivered to customers within accepted standards and in the amount desired." This is more so related to the stochastic or unpredictable variation in the output of a technology. Unlike intermittency, this cannot be planned around with 100\% certainty. A practical example may be a wind turbine's systems failing. While we may know the chances of this occurring, it's not possible to know when it will occur ahead of time; consequently, this may result in a temporary reduction in the quantity and quality of the electricity delivered --  a loss of reliability. In short, when \citeauthor{HH} define renewable output as equal to a baseline capacity multiplied by a uniform random variable, their model is closer to one of reliability rather than intermittency.

On the other hand, we neglect reliability for parsimony and develop a two-technology, two-period model consisting of fossil and renewable energy. Then, we model intermittency as defined above by allowing the output of renewable energy to differ between periods according to known function. At the same time, we consider a representative consumer whose utility is given by CES function where each good is electricity consumption differentiated by time. This captures consumers' preferences for electricity consumption throughout the day as well as their willingness to substitute consumption across time. We then fit the parameters of this CES utility function using electricity consumption and price data for each US state. At the same time, we set parameters of our technologies (their cost and output) to those of coal and solar. Overall, this treatment of intermittency produces some key results similar to that of Helm and Mier. That is, we also find a Pigovian tax to be a solution to externalities along with an S-shaped adoption curve for renewables. But, moreover, we find the elasticity of substitution between fossils and renewables to be highly non-linear; this directly challenges the CES production structure assumption made by standard models of fossil and renewable energy. Finally, we discuss how a combination of carbon taxes and research subsidies may be more optimal than carbon taxes alone in the presence of intermittency.

% What we have in common
% What's new, what adds to the literature

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% Summary of results
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We model electricity as a heterogenous good which is differentiated by its delivery time. Then, we consider a representative consumer who purchases varying quantities of electricity in each period in order to maximize utility. Generally, people prefer to spread their electricity consumption out over time; the convexity of the indifference surfaces of a standard CES function can model this preference in a simple way. The use of a CES function for modeling consumer utility with electricity consumption differentiated by time has been explored earlier by \citet{Moha2016}. Their paper calibrates the share parameters of the CES function with actual electricity consumption data for peak and off-peak hours; they further provide a sensitivity analysis with respect to the elasticity of substitution parameter. Similarly, \citet{Schwarz} provides an empirical study of this CES function and estimate its parameters for industrial customers in a real-time pricing program; their estimates of the intertemporal elasticity of substitution for electricity consumption fall around 0.03.

Next, we model the electricity sector using a representative firm that chooses a set of electricity-generating inputs to maximize profit; some of these inputs are intermittent - they generate a differing amounts of energy in each period. These two sides of the market reach an equilibrium through adjustments in prices of electricity in each period.\footnote{While, in reality, consumers generally pay fixed rates for electricity, \citeauthor{HH} provide a motivation for models incorporating dynamic pricing; they argue such approaches to pricing will become the norm with further technological advances and coming regulatory changes. } For simplicity, we do not allow for load rationing  or negative prices. And, again for simplicity, we assume that the output of all technologies in each period does not vary stochastically.

@@

For more tractable results, we consider a simpler scenario where $n = m = 2$ and $\sigma = 1$ (Cobb-Douglas) which is described in detail in the appendix. To start, this scenario allows us to determine which sets of parameters lead to edge cases where one of the technologies is not used. We may define these conditions in terms of cost efficiency and output efficiency; to clarify, for some arbitrary period $t$ and technology $i$, we use cost efficiency to refer to $\xi_{i,t}/c_i$ and output efficiency to refer to $\xi_{i,t}$. The conditions to avoid an edge case are then given by the following proposition.
\begin{proposition}
	For technology $i$ to be economical, it must meet three conditions with respect to same arbitrary period $t$. Firstly, it must the most cost effective technology for some particular period $t$. Secondly, it must have have a comparative advantage in output efficiency in the same period $t$. And, thirdly, there must be sufficient amount of demand for energy in this period $t$. 
\end{proposition}
The condition on cost efficacy is fairly intuitive; one case of the contrapositive is that if a technology is not cost effective in any period, it will not be used. Similarly, the other is that if a technology is the most cost effective in every period, it will be the only technology used. The second condition regarding output efficiency actually stems from the invertibility of $\xi$. If we did not have $\xi$ invertible, we either have at least one technology that does not produce in any period or we have at least one technology being a linear combination of the other technologies. In the latter case, some technologies are not economical because their output can be replicated the other technologies in a more cost effective way. As an example, suppose $\xi$ consisted of 3 technologies but was of rank $2$. In this case, we may represent any of these technologies as a linear combination of the other two; call this combination the synthetic version of technology $i$. It is not possible for every synthetic technology $i$ to be more expensive than its original; for at least one $i$, we must have a synthetic technology being cheaper or equal in price to the actual technology. In the first case, synthetic technology $i$ is cheaper so we do not use the actual technology $i$. In the later case, we may still eliminate technology $i$ or not use the other two technologies depending on the demand. Overall, in any case, $\xi$ not invertible means at least one technology is not used. Finally, the demand condition is straightforward; even if a technology is optimal in a certain period, if consumers do not sufficiently demand electricity during that period, then there is little reason to use that technology.

We may also derive the comparative statics for this simplified scenario.  
\begin{proposition}
	The equilibrium quantity of a technology is increasing with its output efficiency and decreasing with its cost; at the same time, it is decreasing with the output efficiency of the other technologies and increasing with their cost. Also, suppose that some technology $i$ is the most cost effective in period $t$. Then, its equilibrium quantity is increasing with respect to the demand share parameter $\alpha_t$ and decreasing with respect to the demand share parameters of the other periods. Furthermore, with the same conditions on $i$ and $t$, the comparative statics of $Z_t$ and $X_i$ are equivalent. 
\end{proposition}	
The comparative statics with respect to $X$ and its output efficiency and cost are not surprising. On the other hand, the statics for $Z$ may not be immediately obvious. However, they are still intuitive given that we have $Z = \xi_T X$. That is, suppose that the optimal quantity of $X_1$ rises by 1\% while that of $X_2$ falls by the same amount. Then, we expect output to increase whichever period $\xi_1$ is largest, while it should fall in whichever period $\xi_2$ is smallest; this essentially describes the signs for $\partial Z / \partial X$. Now, since we have already derived the signs of the comparative statics for $X$ with respect to each of the parameters, we can intuit, by the chain rule, the comparative statics for $Z$.\footnote{This intuition would not be correct if the derivative in $X_1$ with respect to an arbitrary parameter was significantly different in magnitude from the derivative in $X_2$ with respect to the same parameter. However, in this particular scenario, such a case does not occur.} Similarly, the comparative statics for the share parameters of the utility function, $\alpha$, follow in the opposite direction. A rise in $\alpha_t$ would directly raise the optimal quantity of $Z_t$; hence, whichever technology is more cost effective at producing in period $t$ would be used more. We provide a more detailed and quantitative discussion of the comparative statics and edge cases in the appendix.

\section{Empirical Methodology}

@@

A number of other papers have approached this problem using data from real-time pricing experiments. These include \citet{Schwarz}, \citet{Herriges}, and \citet{KS1994}.\footnote{There has also been a large literature that directly estimates the price elasticity of electricity demand without imposing a CES functional form. These papers include \citet{Wolak}, \citet{Zarnikau}, \citet{Woo}, \citet{Zhou}, \citet{Reiss}, \citet{Fan}, and \citet{Deryugina}. These papers estimate own-price elasticities, while some also estimate cross-price elasticities for electricity consumption at different times. Because they do not impose a CES structure, we cannot obtain estimates of $\sigma$ from this literature. }
The latter two papers estimate $\sigma$ to be around $0.1$ while the paper by \citeauthor{Schwarz}  obtain estimates around $0.04$. All three papers study real-time electricity pricing programs for industrial consumers using similar methodologies. \citet{Aubin} also provide estimates of the $\sigma$ but using a different methodology; their results find $\sigma < 0$ which is impossible since it would imply upward-sloping demand curves under a CES structure. A common flaw in these papers is that they do not account for endogeneity caused by a supply-side response to prices. For instance, coal plants are able to modify their generation between days; so, they might respond to an increase in future prices by storing coal until prices rise. That is, there may exist intertemporal substitution for electricity generation which would bias $\sigma$ downward unless controlled for. This is particularly important, because whether $\sigma$ is closer to 0.1 or 1 can make a significant difference with respect to the practical implications of our model.

@@

We collect monthly data from the \cite{EIANetgen} on retail electricity prices and consumption for each state in the US from 2011 to 2018. Additionally, also from the EIA, we obtain data on the average cost of coal for electricity generation for each state and month.\footnote{The coal price data set contains a large number of missing values due to privacy concerns; however, we do not expect that these missing values are correlated with the data itself.} We deflated both electricity and coal prices over time using the PCECP Index provided by the  \cite{USBEA}. 
Finally, we collect data on HDDs and CDDs from the NOAA's \citet{CPC} for the same panel. Then, we merge these three data sets and trim 1\% of outliers for a total of $826$ observations for each month and state. We use this preliminary data set to construct the data required for our regressions. That is, each observation in our estimation equation belongs to a set $(t,s,i)$ consisting of two time periods and a state. Hence, we construct each row in our regression data set using unique combinations of $t,s$ where $t \neq s$ for each state $i$. Due to the number of potential combinations and thus observations, we restrict our data set to a random sample of 9000 observations.\footnote{We reran our regressions with several different samples of 9000 and found that our estimated coefficients did not significantly change; hence we believe this sample size is sufficient.} 
All in all, each observation in our regression data set consists the following variables: state ($i$), date 1 ($t$), date 2 ($s$), the log difference in electricity consumption between month 1 and month 2, the log difference in the price of electricity, the log difference in the price of coal, the number of CDDs for each date, the number of HDDs for each date, and the difference (in months) between dates 1 and 2.

@@

Firstly, we let technology 1 be coal power and technology 2 be solar power; again, period $t$ represents peak hours and period $s$ represents off-peak hours. We assume that, holding prices equal, consumers prefer that 60\% of their energy arrive in period $t$ and the remaining 40\% arrive in period $s$; that is, we have $\alpha_t = 0.6$ and $\alpha_s = 0.4$. Additionally, we normalize the units of the quantities to a MWh basis. Hence, $\xi$ represents the percent of capacity utilized in each period; we assume coal uses 100\% of its capacity in both periods, while solar can access 100\% during peak hours and only 10\% during the off-peak. Finally, we set cost parameters (given in  \$ per MWh)  equal to LCOE estimates for 2023 from the \citet{EIALCOE}; specifically, we use estimates for "Solar PV" and "Coal with 30\% CCS" from Table 1b. In total, we have the following parameters.

@@

\subsection{Cobb-Douglas Case with Two Periods \& Two Technologies}

@@

>>>>

@@

Renewable energy technologies have seen considerable adoption over the last few decades (EIA 2019). Unlike the alternatives, wind and solar power are particularly unique in that the amount of energy they supply is intermittent. Consequently, designing economically efficient policy to promote their adoption is not straightforward given that they cannot easily substitute for fossil fuel technologies such as coal power. Moreover, traditional economic measures such as the levelized cost of electricity (LCOE) fail to capture the true economic value of intermittent technologies, because they neglect to account for variation in output and prices over time \citep{Joskow2011}.

Some of the literature has overcome this critique by constructing numerical models that find the cheapest renewable technology set while accounting for intermittent supply. For instance, \citet{MN2006} model uncertain renewable output with intertemporal generation constraints, while \citet{NCK2007} model temporal and spatial characteristics of wind output to optimize its deployment in the UK. On the other hand, other literature focuses on the effect of intermittent technologies on the market itself; \citet{AC2012} study the interaction between intermittent renewables and traditional reliable sources of energy in decentralized markets, and \citet{Chao2011} models alternative pricing mechanisms for intermittent renewable energy sources. Additionally, \citet{Boren2012} reviews the effects of present public policies used to promote renewables and the challenges posed by intermittency.

Our model comes closest to that of \citet{HH} who build a peak-load pricing model where the availability of renewable capacity varies stochastically. This stochastic variability models intermittency and negatively affects the adoption of renewables. Furthermore, they simplify their analysis by setting aside complications such as  outage costs and rationing rules. In total, their model finds an S-shaped adoption curve for renewables as they get cheaper. In addition, they find that a Pigouvian tax can properly internalize the costs of fossil fuels in a setting with perfect competition. Likewise, we model the  equilibrium of a market with dynamic prices and access to both renewable and fossil energy. However, our results differ in key ways, because we take a different approach to modeling intermittency.

To highlight this difference, we must first contrast the terms  \textit{intermittency} and \textit{reliability}. By intermittency, we predictable changes in output related to physical constraints. For instance, while wind energy output varies over time, if  we know the angle and speed of wind, we may precisely derive the quantity of energy that a wind farm generates.\footnote{ \citet{Foley2012} provide a review of the literature on the forecasting of wind power. Over time, forecasts have gotten much more accurate allowing electricity grids to manage wind power intermittency ahead of time.}  At the same time, these forecasts may not be perfect and, as a result, we see unexpected variation in the output of a technology. This unpredictability may be better captured by the notion of reliability; specifically,  we refer to definition provided by the US Department of Energy's \citet{ORNL} -- ``Power reliability can be defined as the degree to which the performance of the elements in a bulk system results in electricity being delivered to customers within accepted standards and in the amount desired." This is more so related to the stochastic or unpredictable variation in the output of a technology. Unlike intermittency, this cannot be planned around with 100\% certainty. A practical example may be a wind turbine's systems failing. While we may know the chances of this occurring, it's not possible to know when it will occur ahead of time; consequently, this may result in a temporary reduction in the quantity and quality of the electricity delivered --  a loss of reliability. In short, when \citeauthor{HH} define renewable output as equal to a baseline capacity multiplied by a uniform random variable, their model is closer to one of reliability rather than intermittency.

On the other hand,  we model intermittency as defined above by allowing the output of renewable energy to differ between periods according to known function. But, for parsimony, we assume that the output of all technologies in each period does not vary stochastically; that is, we focus on intermittency rather than reliability. We then model the electricity sector using a representative firm that chooses and builds capacity from a set of electricity-generating technologies to maximize profit; some of these technologies are intermittent while others have constant output. Then, we consider a representative consumer who purchases varying quantities of electricity in each period in order to maximize utility. Generally, people prefer to spread their electricity consumption out over time; a standard CES function can capture this preference in a simple way. That is, we treat electricity as a heterogenous good which is differentiated by its delivery time; consumer utility is given by a CES function of electricity consumption in each period.\footnote{ The use of a CES function as such has been explored earlier by \citet{Schwarz}, \citet{Schwarz}, \citet{Herriges}, and \citet{KS1994}, \citet{Aubin}, and \citet{Moha2016}. Their papers empirically estimate the parameters for this function; we provide a more detailed discussion of the empirical literature in our \hyperref[sec:methodology]{Methodology}.} These two sides of the market reach an equilibrium through adjustments in prices of electricity in each period. All in all, we make many of the same assumptions that \citet{HH} do. We assume dynamic pricing,\footnote{ \citeauthor{HH} provide a motivation for models incorporating the dynamic pricing of electricity. They argue such approaches to pricing will become the norm with further technological advances and coming regulatory changes.} no load rationing, and positive prices.

Then, we parametrize our model empirically. To start, we fit the parameters of the consumer's CES utility function using electricity consumption and price data for each US state. Specifically, we estimate the intertemporal elasticity of substitution for electricity consumption; this parameter plays a particularly important role in our model, since it captures the effects of intermittency on demand. Then, to model the supply side, we narrow our model to a two-period, two-technology setting with renewables and fossil energy. We proxy for renewable energy using solar and proxy for fossil energy using coal; then, we parametrize each accordingly. Finally, we implement our model numerically and solve for the equilibrium. We study the effects of technical change on the energy market and derive policy implications.

Overall, this treatment of intermittency produces some key results that contrast with those of Helm and Mier. First, while we also find that externalities can be handled through a Pigouvian tax, we further find that a combining both taxes on pollution and research subsidies for renewables is more optimal than either policy instrument alone. Second, we find that, as renewables get cheaper, their adoption is not necessarily S-shaped but can be concave or convex depending the elasticity of substitution in the consumer's utility function; with our estimated parameters, we find this relationship to be almost linear. Third, a significant amount of literature has assumed a CES structure between renewable and fossil energy (see \citet{Pap}); this assumption has been motivated, in part, by the need to capture imperfect substitutability between these two generation technologies as a result of intermittency. However, our direct model of intermittency disconfirms this assumption; the elasticity of substitution between these two energy technologies is far from linear as a \textit{consequence} of intermittency. In short, optimal taxes and subsidies along with the substitutability of renewable and fossil energy all vary with the state of the local energy market. Hence, we believe that optimal environmental policy must take into account local conditions far more than has previously been assumed.

@@

@@

For more tractable results, we consider a simpler scenario where $n = m = 2$ and $\sigma = 1$ (Cobb-Douglas); this particular case is described in greater  detail with proofs in \hyperref[sec:cobbdoug]{Appendix A.A}. Moreover, these parameters are of particular interest, because they simplify the model enough to allow us to derive comparative statics and determine which sets of parameters lead to edge cases.

To start, we discuss the conditions on the exogenous parameters required to avoid edge-cases. Alternatively, these are the conditions required for a technology to be economical. But, first, we define two recurring terms in our analysis: cost efficiency and output efficiency. For some arbitrary period $t$ and technology $i$, we use cost efficiency to refer to $\xi_{i,t}/c_i$ and output efficiency to refer to $\xi_{i,t}$. So, for example, A technology is more cost efficient than another in a particular period when its output in that period per ``dollar" spent is larger than that of the other technology in the same period. 
\begin{proposition}
	Assume that, for all $i$ and $t$, we have $\xi_{i,t}, \alpha_t, c_i > 0$. Furthermore, suppose that $\xi$ is invertible. For technology $i$ to be economical, it must meet three conditions with respect to some arbitrary period $t$. Firstly, it must the more cost effective than any of the other technologies in period $t$. Secondly, it must have have a comparative advantage in output efficiency in the same period $t$. And, thirdly, there must be sufficient amount of demand for energy in this period $t$. 
\end{proposition}
The condition on cost efficacy is fairly intuitive; one case of the contrapositive is that if a technology is not the most cost effective in any period, it will not be used. Alternatively, if a technology is the most cost effective in every single period, it will be the only technology used. The second condition regarding output efficiency actually stems from the invertibility of $\xi$. If we did not have $\xi$ invertible, we either have at least one technology that does not produce in any period or we have at least one technology being a linear combination of the other technologies in terms of output. In the latter case, some technologies are not economical because their output can be replicated the other technologies in a more cost effective way. As an example, suppose $\xi$ consisted of 3 technologies but was of rank $2$. In this case, we may represent any of these technologies as a linear combination of the other two; call this combination the synthetic version of technology $i$. It is not possible for every synthetic technology $i$ to be more expensive than its original; for at least one $i$, we must have a synthetic technology being cheaper or equal in price to the actual technology. In the first case, synthetic technology $i$ is cheaper so we do not use the actual technology $i$. In the later case, we may still eliminate technology $i$ or not use the other two technologies depending on the demand. Overall, in any case, $\xi$ not invertible means that at least one technology is not used. Finally, the demand condition is straightforward; even if a technology is optimal in a certain period, if consumers do not sufficiently demand electricity during that period, then there is little reason to use that technology.

We may also derive the comparative statics for this simplified scenario.  
\begin{proposition}
	Suppose we are not in an edge case, so that the conditions of Proposition 1 hold for each technology. The equilibrium quantity of a technology is increasing with its output efficiency and decreasing with its cost; at the same time, it is decreasing with the output efficiency of the other technologies and increasing with their cost. Also, suppose that some technology $i$ is the most cost effective in period $t$. Then, its equilibrium quantity is increasing with respect to the demand share parameter $\alpha_t$ and decreasing with respect to the demand share parameters of the other periods. Furthermore, again assuming technology $i$ is the most cost effective in period $t$, the comparative statics of $Z_t$ and $X_i$ are equivalent. 
\end{proposition}	
The comparative statics with respect to $X$ and its output efficiency and cost are not surprising. On the other hand, the statics for $Z$ may not be immediately obvious. Instead, they follow from the fact that we have $Z \equiv \xi_T X$. That is, suppose we have an arbitary technology $i$ that is the most cost effective source of electricity on period $t$. If consumers demanded that 100\% of their energy arrive in period $t$, then relying solely on technology $i$ for energy would be the most economical solution. Consequently, it seems intuitive that the comparative statics of $X_i$ follow through to $Z_t$. This intuition just happens to follow through even when other technologies are employed and there is demand in multiple periods. Similarly, the comparative statics for the share parameters of the utility function, $\alpha$, travel in the opposite direction. A rise in $\alpha_t$ would directly raise the optimal quantity of $Z_t$; hence, whichever technology is most cost effective at producing in period $t$ would be used more. We provide a more detailed and quantitative discussion of the comparative statics and edge cases in the appendix.

\section{Empirical Methodology}
\label{sec:methodology}

@@

A number of other papers have approached this problem using data from real-time pricing experiments. These include \citet{Schwarz}, \citet{Herriges}, and \citet{KS1994}.\footnote{ There has also been a large literature that directly estimates the price elasticity of electricity demand without imposing a CES functional form. These papers include \citet{Wolak}, \citet{Zarnikau}, \citet{Woo}, \citet{Zhou}, \citet{Reiss}, \citet{Fan}, and \citet{Deryugina}. These papers estimate own-price elasticities, while some also estimate cross-price elasticities for electricity consumption at different times. Because they do not impose a CES structure, we cannot obtain estimates of $\sigma$ from this literature. }
The latter two papers estimate $\sigma$ to be around $0.1$ while the paper by \citeauthor{Schwarz}  obtain estimates around $0.04$. All three papers study real-time electricity pricing programs for industrial consumers using similar methodologies. Additionally, \citet{Aubin} also provide estimates of the $\sigma$ but using a different methodology; their results find $\sigma < 0$ which is impossible since it would imply upward-sloping demand curves under a CES structure. Finally, \citet{Moha2016} also empirically estimates the share parameters for a CES function of this form, but do not estimate the elasticity of substitution.

A common flaw in these papers is that they do not account for endogeneity caused by a supply-side response to prices. For instance, coal plants are able to modify their generation between days; so, they might respond to an increase in future prices by storing coal until prices rise. That is, there may exist intertemporal substitution for electricity generation which would bias $\sigma$ downward unless controlled for. This is particularly important, because whether $\sigma$ is closer to 0.1 or 1 can make a significant difference with respect to the practical implications of our model.

@@

We collect monthly data from the \cite{EIANetgen} on retail electricity prices and consumption for each state in the US from 2011 to 2018. Additionally, also from the EIA, we obtain data on the average cost of coal for electricity generation for each state and month.\footnote{ The coal price data set contains a large number of missing values due to privacy concerns; however, we do not expect that these missing values are correlated with the data itself.} We deflated both electricity and coal prices over time using the PCECP Index provided by the  \cite{USBEA}. 
Finally, we collect data on HDDs and CDDs from the NOAA's \citet{CPC} for the same panel. Then, we merge these three data sets and trim 1\% of outliers for a total of $826$ observations for each month and state. We use this preliminary data set to construct the data required for our regressions. That is, each observation in our estimation equation belongs to a set $(t,s,i)$ consisting of two time periods and a state. Hence, we construct each row in our regression data set using unique combinations of $t,s$ where $t \neq s$ for each state $i$. Due to the number of potential combinations and thus observations, we restrict our data set to a random sample of 9000 observations.\footnote{ We reran our regressions with several different samples of 9000 and found that our estimated coefficients did not significantly change; hence we believe this sample size is sufficient.} 
All in all, each observation in our regression data set consists the following variables: state ($i$), date 1 ($t$), date 2 ($s$), the log difference in electricity consumption between month 1 and month 2, the log difference in the price of electricity, the log difference in the price of coal, the number of CDDs for each date, the number of HDDs for each date, and the difference (in months) between dates 1 and 2.

@@

Firstly, we let technology 1 be coal power and technology 2 be solar power; again, period $t$ represents peak hours and period $s$ represents off-peak hours. We assume that, holding prices equal, consumers prefer that 60\% of their energy arrive in period $t$ and the remaining 40\% arrive in period $s$; that is, we have $\alpha_t = 0.6$ and $\alpha_s = 0.4$. Additionally, we normalize the units of the quantities to a MWh basis. Hence, $\xi$ represents the percent of capacity utilized in each period; we assume coal uses 100\% of its capacity in both periods, while solar can access 100\% during peak hours and only 10\% during the off-peak. Finally, we set cost parameters (given in  \$ per MWh)  equal to LCOE estimates for 2023 from the \citet{EIALCOE}; specifically, we use estimates for "Solar PV" and "Coal with 30\% CCS" from Table 1b.\footnote{ The LCOE for a technology is equal to the sum of its lifetime costs divided by its lifetime energy output. While \citet{Joskow2011} explains the flaws of comparing generation technologies solely on the basis of LCOE, our use of this measure is unrelated to his critique. That is, he argues that the economic value an intermittent technology should account for when it produces electricity and the prices of electricity in those periods (see Table 2 of his paper). Our model does exactly this.}  In total, we have the following parameters.

@@

\subsection{Cobb-Douglas Case with Two Periods \& Two Technologies}
\label{sec:cobbdoug}

@@

	\bibitem[Papageorgiou et al.(2017)]{Pap}
	Papageorgiou, C., Saarn, M., \& Schulte, P. (2017). Substitution between clean and dirty energy inputs: a macroeconomic perspective.(Author abstract). Review of Economics and Statistics, 99(2), 201–212. https://doi.org/10.1162/REST\_a\_00623
